---
title: "Webinar 1 - R for Data Management"
author: "Frank Namugera (PhD)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
    highlight: tango

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

This R Markdown file is the **live-coding companion** to Webinar 1. It follows
the session agenda below and uses a real **Malawi Demographic and Health
Survey (MDHS)** Individual Recode (women's) file throughout, so every function
is demonstrated on genuine survey data rather than a toy example.

## Session Setup
### 1. Install R

Download R from:

https://cran.r-project.org/

Steps:
- Select your operating system (Windows/macOS/Linux).
- Download and install the latest version.
- Accept the default installation settings.

---

### 2. Install RStudio

Download RStudio Desktop from:

https://posit.co/download/rstudio-desktop/

Steps:
- Install RStudio after installing R.
- Open RStudio and confirm that R is detected.

## Research Focus

**Determinants of Modern Contraceptive Use among Women Aged 15--49 Years**

Modern contraceptive use is a key indicator of reproductive health and is
essential for reducing unintended pregnancies, improving maternal and child
health outcomes, and supporting sustainable population growth. Using data
from the Malawi DHS, this session examines the determinants of modern
contraceptive use among women aged 15--49, and how it varies by age,
education, marital status, wealth, place of residence, region, employment,
media exposure, and number of living children.

If you have downloaded your own DHS extract (`.dta` or `.sav`) from the
[DHS Program website](https://dhsprogram.com/data/available-datasets.cfm),
point `file_path` in the import step below to it.

**Tip:** DHS data requires free registration and project approval before
download -- start that process early if you plan to use your own country's
data.

---

# R & RStudio Basics

## How R Works

R has a set of methods and functions designed and packaged into libraries.
For every programming task, you need to know the appropriate functions and
packages to install and use; where a ready-made function doesn't exist, you
can write your own (a "user-defined function").

RStudio is the IDE used to manage your datasets, scripts, and visualizations
in one place.

## Creating and Managing an R Project

### Step 1: Create a New Project

- Open **RStudio**
- Go to **File -> New Project**
- Select **New Directory -> New Project**
- Enter a project name and choose a location
- Click **Create Project**

### Step 2: Organize Your Project

Create folders to keep your work organized:

```
Project_Name/
|
|-- data/
|-- scripts/
|-- figures/
|-- output/
|-- reports/
`-- Project_Name.Rproj
```

## Good Practices

- Use one project per analysis
- Organize files into folders
- Use relative file paths instead of hard-coded `setwd()` calls
- Save your work regularly
- Keep raw and cleaned data separate

## Basic R Operations

R can handle everything from simple arithmetic to full statistical models.
We will build up to the DHS analysis gradually, starting with the basics:

```{r assignment}
## the assignment operator
a <- 2
b <- 5:10
c <- c("a", "b", "d")
```

## Loading Required Packages

```{r packages}
# install.packages(c("tidyverse", "haven", "readxl", "labelled"))

library(tidyverse)   # dplyr, ggplot2, tidyr, stringr, etc.
library(haven)       # read Stata (.dta) / SPSS (.sav) files -- standard DHS formats
library(readxl)      # read Excel files
library(labelled)    # helper functions for labelled survey data
```

```{r working-directory}
# Check and, if needed, set your working directory
getwd()
# setwd("path/to/your/project")
```

---

# Importing and Inspecting Data

## Built in datasets

R comes with several datasets under the different libraries. Sometimes the research questions could be answered by some of these datasets.

```{r inbuilt_data}
# This functions shows you all available datasets under the base function

# data()

```

## Supported Data Formats

R can read data from many formats, including `.csv`, `.xlsx`, `.sav`, `.dta`,
`.R`/`.RData`, and more. This session focuses on `.dta` (Stata format), since
that is how the Malawi DHS Individual Recode file is distributed.

## Importing the DHS Dataset

DHS data comes **already value-labelled**: every categorical variable (e.g.,
`v025` place of residence, `v106` education) carries its numeric code *and*
a text label at the same time, via haven's `labelled` class. We keep both
together rather than converting everything to plain factors on import,
that way we still have the numeric codes available for `na_if()` and other
numeric checks, but can pull in the ready-made labels with `as_factor()`
whenever we want them (shown in **Working with Categorical Variables**
below).

```{r import-real}
# Update this path to point to your own DHS extract
# change these paths to make reference to the directory that has the dataset.
file_path_dta <- "C:/Users/FRANK/Desktop/R training/DHS_data.dta" 
file_path_csv<- "C:/Users/FRANK/Desktop/R training/DHS_data.csv"

dhs_data <- read_dta(file_path_dta) 
dhs_data<- mutate(dhs_data, across(where(is.labelled), as_factor)) 

## piping function
dhs_data <- read_dta(file_path_dta) |>
  mutate(across(where(is.labelled), as_factor))  ### haven and convert values to value labels.

```

## Selecting Variables of Interest

```{r variable-selection}
var_names <- c(
  "caseid", # case id
  "v005",   # sampling weight
  "v021",   # primary sampling unit (cluster)
  "v022",   # sampling strata
  "v024",   # region
  "v025",   # place of residence
  "v012",   # age
  "v106",   # highest education level
  "v130",   # religion
  "v190",   # wealth index
  "v201",   # children ever born
  "v218",   # number of living children
  "v213",   # currently pregnant
  "v312",   # current contraceptive method
  "v313",   # current contraceptive method type
  "v501",   # marital status
  "v714",   # currently working
  "v717",   # respondent's occupation
  "v701",   # husband's education
  "v705",   # husband's occupation
  "v157",   # reads newspaper
  "v158",   # listens to radio
  "v159",   # watches television
  "v394",   # visited health facility
  "v481"    # health insurance coverage
  )

dhs_raw <- dhs_data %>% select(var_names)

# other aproaches such as dhs_data[c(var_names)]
```

`v005`, `v021`, and `v022` (the sampling weight, cluster, and strata) are not
used in this session's basic descriptive analysis, but we keep them selected
now, they're needed for **survey-weighted analysis**, which is previewed in Webinar 2.

## First Look at the Data

```{r inspect}
dim(dhs_raw);nrow(dhs_raw);ncol(dhs_raw)

# overview
head(dhs_raw)


tail(dhs_raw)


names(dhs_raw)


## data types
str(dhs_raw)


glimpse(dhs_raw)

## data summary/descriptives
summary(dhs_raw)
```


# Data Cleaning -- Part 1

## Objectives

By the end of this segment, participants should be able to:

- Rename variables
- Identify and handle missing values
- Detect and remove duplicate records
- Recode variables
- Work with categorical variables
- Detect and handle outliers

## Renaming Variables

DHS datasets use standardized variable names (e.g., `v012`, `v106`). Renaming
them up front makes every subsequent step easier to read.

```{r rename}
# shortcut for %>%  is ctr+shift+m %>% 

dhs <- dhs_raw |>
  rename(
    age                     = v012,
    region                  = v024,
    residence               = v025,
    education               = v106,
    religion                = v130,
    wealth_quintile         = v190,
    children_ever_born      = v201,
    pregnant                = v213,
    living_children         = v218,
    contraceptive           = v313,
    marital_status          = v501,
    currently_working       = v714,
    occupation              = v717,
    husband_education       = v701,
    husband_occupation      = v705,
    reads_newspaper         = v157,
    listens_radio           = v158,
    watches_tv              = v159,
    visited_health_facility = v394,
    health_insurance        = v481
  )

names(dhs)
```

## Identifying Missing Values

```{r missing-check}
# Count missing values for every column
# in a single variable:
sum(is.na(dhs$marital_status))

## overall:
colSums(is.na(dhs))
```

## Handling Missing Values

DHS commonly uses special numeric codes for "missing" instead of
a blank cell. We convert these to proper `NA` so R treats them correctly.

```{r missing-recode}

colSums(is.na(dhs))

# Let's check the husband_education & husband_occupation

# dhs %>% filter(is.na(husband_education)) |> dim

### sometimes missingness is a design intent not an error. What can you seen in common in all these women?

# dhs %>% filter(is.na(husband_education)) |> view() 

# Muted since there are no missing ness.
# dhs <- dhs |>
#   mutate(
#     age             = na_if(age, 98),
#     education       = na_if(education, 9),
#     pregnant        = na_if(pregnant, 9),
#     living_children = na_if(living_children, 99)
#   )

```

## Removing Duplicate Records

Duplicate records can give a misleading picture of the data by over-weighting
a small number of respondents.

```{r duplicates}
# How many caseids appear more than once?
## use the duplicated() function

sum(duplicated(dhs$caseid))

# Inspect the duplicated rows themselves (both occurrences)
dhs_duplicates <- dhs |>
  filter(duplicated(caseid) | duplicated(caseid, fromLast = TRUE))


dim(dhs_duplicates)
dhs_duplicates

# Keep only one unique occurrence of each caseid

dhs <- dhs |> distinct(caseid, .keep_all = TRUE)
nrow(dhs)
```

## Recoding Variables

`residence` and `education` are already value-labelled by DHS, so we don't
need to manually retype their coding scheme -- that comes in the next
section via `as_factor()`. The one variable that genuinely needs manual
recoding here is `age_group`, since it's a brand-new category that doesn't
exist as a raw DHS variable and so has no label to inherit:

```{r recode}
dhs <- dhs |>
  mutate(
    age_group = case_when(
      age >= 15 & age < 20  ~ "15-19",
      age >= 20 & age < 35  ~ "20-34",
      age >= 35 & age <= 49 ~ "35-49",
      TRUE ~ NA_character_
    )
    
    # An equivalent (but harder to read for 3+ categories) version with
    # nested ifelse() would be:
    # age_group = ifelse(age >= 15 & age < 20, "15-19",
    #             ifelse(age >= 20 & age < 35, "20-34",
    #             ifelse(age >= 35 & age <= 49, "35-49", NA)))
  )

count(dhs, age_group)
```

## Working with Categorical Variables (Factors)

Because `residence` and `education` are already value-labelled, `as_factor()`
pulls in DHS's own text labels directly, no manual mapping required, and no
risk of a typo creating a mismatched category:

```{r factors}
dhs <- dhs |>
  mutate(
    residence_f = as_factor(residence),
    education_f = factor(as_factor(education), ordered = TRUE),
    age_group_f = factor(age_group, levels = c("15-19", "20-34", "35-49"))
  )

levels(dhs$residence_f)
levels(dhs$education_f)
table(dhs$education_f, useNA = "ifany")
```

`as_factor()` gives us the categories DHS already defined. Manual recoding
still earns its keep when you want *different* groupings than the source
labels provide -- for example, collapsing four education levels down to two
for a simpler comparison:

```{r factors-collapse}
dhs <- dhs |>
  mutate(
    education_grouped = fct_collapse(
      education_f,
      "No education"      = "No education",
      "Primary or higher" = c("Primary", "Secondary", "Higher")
    )
  )

table(dhs$education_grouped)
```

## Outlier Detection and Removal

Outliers are values that fall far outside a variable's typical range. Left
unaddressed, they can distort means, standard deviations, and any models
built on the data. Two complementary checks are useful:

1. **Validity / range checks**: for variables with a known plausible range.
   This survey only interviews women aged 15--49, so any value outside that
   range is a data error, not a genuine outlier.
2. **Statistical outlier detection**: for continuous variables without a
   fixed valid range (e.g., number of living children), where a rule such as
   the interquartile range (IQR) flags unusually extreme values.

### Validity check: age

```{r validity-check}
# This should return 0 rows if age was recorded correctly
dhs %>% filter(age < 15 | age > 49)
```

### Statistical outlier detection: living children

```{r outlier-boxplot, fig.width=5, fig.height=4}
boxplot(dhs$living_children,
        main = "Number of Living Children",
        ylab = "Living children",
        col  = "lightblue")
```

```{r outlier-iqr}
Q1      <- quantile(dhs$living_children, 0.25, na.rm = TRUE)
Q3      <- quantile(dhs$living_children, 0.75, na.rm = TRUE)
IQR_val <- IQR(dhs$living_children, na.rm = TRUE)

lower_bound <- Q1 - 1.5 * IQR_val
upper_bound <- Q3 + 1.5 * IQR_val

c(lower_bound = lower_bound, upper_bound = upper_bound)

# How many observations fall outside these bounds?
dhs %>%
  filter(living_children < lower_bound | living_children > upper_bound) %>%
  nrow()
```

```{r outlier-flag-remove}
# Flag outliers rather than silently deleting them, so the decision stays
# auditable
dhs <- dhs |>
  mutate(
    living_children_outlier = living_children < lower_bound |
                               living_children > upper_bound
  )

table(dhs$living_children_outlier, useNA = "ifany")

# Remove confirmed outliers (missing values were already handled earlier,
# so we keep NAs rather than dropping them here)
dhs <- dhs |>
  filter(is.na(living_children) | !living_children_outlier)
```

> **Note:** removing outliers is not always the right choice -- a woman may
> genuinely have an unusually high number of children. Always inspect
> flagged cases before deleting them, and document the rule you used.

---

# Data Cleaning -- Part 2

## Objectives

Participants should be able to:

- Create new variables using `mutate()`
- Select variables using `select()`
- Filter observations using `filter()`
- Arrange data using `arrange()`
- Combine cleaning steps into a single pipeline

## Creating New Variables with `mutate()`

The DHS variable `contraceptive` (originally `v313`) classifies contraceptive
methods as:

| Code | Method Type |
|---|---|
| 0 | No method |
| 1 | Folkloric |
| 2 | Traditional |
| 3 | Modern |

```{r mutate-modern-use}
dhs <- dhs |>
  mutate(modern_use = ifelse(contraceptive == "modern method", "yes", "no"))

head(dhs$modern_use)
```

We also create a media exposure indicator for women exposed to at least one
media source:

```{r mutate-media}
dhs <- dhs |>
  mutate(
    media_exposure = ifelse(
      reads_newspaper != "not at all"  | listens_radio != "not at all" | watches_tv != "not at all",
      "Yes", "No"
    )
  )

table(dhs$media_exposure, useNA = "ifany")
```

## Selecting Variables with `select()`

General syntax:

```r
data |> select(variable1, variable2)
```

**Select a few variables to preview:**

```{r select-preview}
dhs |>
  select(age, education_f, residence_f, modern_use) %>%
  head()
```

**Remove variables you no longer need** (e.g., the individual media items,
once `media_exposure` has been derived from them):

```{r select-remove}
dhs %>%
  select(-reads_newspaper, -listens_radio, -watches_tv) %>%
  names()
```

**Select by name pattern** -- most useful *before* renaming, directly on the
raw DHS codes:

```{r select-pattern}
# Variables starting with "v1"
dhs_raw %>% select(starts_with("v1")) %>% names()

# Variables ending in "01"
dhs_raw %>% select(ends_with("01")) %>% names()
```

## Filtering Observations with `filter()`

General syntax:

```r
data %>% filter(condition)
```

```{r filter-examples}
# Women aged 15-49
women_15_49 <- dhs %>% filter(age >= 15 & age <= 49)

# Rural women
rural <- dhs %>% filter(residence == "rural")

# Urban women
urban <- dhs %>% filter(residence == "urban")

# Married women
married <- dhs %>% filter(marital_status == "married")

# Women using a modern contraceptive method
users <- dhs %>% filter(modern_use == "yes")

# Multiple conditions at once: rural, secondary+ education, modern users
subset_data <- dhs %>%
  filter(residence == 'ubarn', education %in%  c("secondary", "higher"), modern_use == "yes")

nrow(women_15_49); nrow(rural); nrow(urban); nrow(married); nrow(users); nrow(subset_data)
```

## Arranging Data with `arrange()`

```{r arrange-examples}
# Youngest to oldest
dhs %>% arrange(age) %>% select(caseid, age) %>% head()

# Oldest to youngest
dhs %>% arrange(desc(age)) %>% select(caseid, age) %>% head()
```

## Putting It All Together

The steps above were shown one at a time for teaching purposes. In practice,
you would chain them into a single reproducible pipeline. This recap starts
again from the **raw, unrenamed** variable selection (`dhs_raw`) and produces
the final analysis dataset we will use for the rest of this session -- reusing
the same outlier bounds computed earlier:

```{r full-pipeline}
analysis_data <- dhs_raw %>%
  rename(
    age                = v012,
    region             = v024,
    residence          = v025,
    education          = v106,
    religion           = v130,
    wealth_quintile    = v190,
    children_ever_born = v201,
    pregnant           = v213,
    living_children    = v218,
    contraceptive      = v313,
    marital_status     = v501,
    reads_newspaper    = v157,
    listens_radio      = v158,
    watches_tv         = v159
  ) %>%
  distinct(caseid, .keep_all = TRUE) %>%
  mutate(
    residence_f = as_factor(residence),
    education_f = factor(as_factor(education), ordered = TRUE),
    age_group_f = factor(case_when(
                    age >= 15 & age < 20  ~ "15-19",
                    age >= 20 & age < 35  ~ "20-34",
                    age >= 35 & age <= 49 ~ "35-49",
                    TRUE ~ NA_character_),
                    levels = c("15-19", "20-34", "35-49")),
    modern_use     = ifelse(contraceptive == "modern method","yes", "no"),
    media_exposure = ifelse(reads_newspaper != "not at all"  | listens_radio != "not at all" | watches_tv != "not at all", "Yes", "No")) %>%
  filter(is.na(living_children) |
           !(living_children < lower_bound | living_children > upper_bound)) %>%
  filter(age >= 15 & age <= 49) %>%
  select(caseid, age, age_group_f, residence_f, education_f, wealth_quintile,
         region, marital_status, children_ever_born, living_children,
         modern_use, media_exposure)

glimpse(analysis_data)
```

`analysis_data` is now our single, clean, analysis-ready dataset. Every
descriptive statistic from here on uses it.

---

# Descriptive Statistics

## Objectives

Participants should be able to:

- Generate frequencies
- Calculate means and medians
- Produce percentages
- Create summary tables

## Numerical Summaries

```{r numerical-summaries}
mean(analysis_data$age, na.rm = TRUE)
median(analysis_data$age, na.rm = TRUE)
range(analysis_data$age, na.rm = TRUE)
sd(analysis_data$age, na.rm = TRUE)
quantile(analysis_data$age, na.rm = TRUE)
summary(analysis_data$age)
```

## Frequency Tables

```{r frequency-tables}
table(analysis_data$education_f)
table(analysis_data$residence_f)
table(analysis_data$modern_use)
```

## Relative Frequencies and Percentages

```{r percentages}
prop.table(table(analysis_data$education_f))

round(prop.table(table(analysis_data$education_f)) * 100, 1)
```

## Two-Way Tables

```{r two-way}
# Education by contraceptive use
table(analysis_data$education_f, analysis_data$modern_use)

# Row percentages (within each education level)
round(prop.table(table(analysis_data$education_f,
                        analysis_data$modern_use), 1) * 100, 1)

# Column percentages (within each contraceptive-use group)
round(prop.table(table(analysis_data$education_f,
                        analysis_data$modern_use), 2) * 100, 1)
```

## Descriptive Statistics Using `dplyr`

```{r dplyr-summary}
analysis_data %>%
  summarise(
    Mean_Age   = mean(age, na.rm = TRUE),
    Median_Age = median(age, na.rm = TRUE),
    SD_Age     = sd(age, na.rm = TRUE),
    Minimum    = min(age, na.rm = TRUE),
    Maximum    = max(age, na.rm = TRUE)
  )
```

## Grouped Summaries

```{r grouped-residence}
analysis_data %>%
  group_by(residence_f) %>%
  summarise(
    Women      = n(),
    Mean_Age   = mean(age, na.rm = TRUE),
    Median_Age = median(age, na.rm = TRUE)
  )
```

```{r grouped-education}
analysis_data %>%
  group_by(education_f) %>%
  summarise(
    Women      = n(),
    Mean_Age   = mean(age, na.rm = TRUE),
    Modern_Use = mean(modern_use, na.rm = TRUE) * 100
  )
```

```{r grouped-multi}
analysis_data %>%
  group_by(residence_f, education_f) %>%
  summarise(
    Women      = n(),
    Mean_Age   = mean(age, na.rm = TRUE),
    Modern_Use = mean(modern_use, na.rm = TRUE) * 100,
    .groups = "drop"
  )
```

## Simple Summary Tables for Reporting

```{r simple-summary-tables}
analysis_data %>% count(residence_f)
analysis_data %>% count(education_f)
analysis_data %>% count(residence_f, modern_use)
```

---

# Practical Exercise

Using the Malawi DHS dataset:

1. Import the dataset and select the variables of interest.
2. Rename variables to meaningful names.
3. Identify and handle missing values.
4. Check for and remove duplicate records.
5. Recode residence, education, and age into meaningful categories, and
   convert them to factors.
6. Detect and handle outliers in the number of living children using the
   IQR method.
7. Create a binary variable for modern contraceptive use and a media
   exposure indicator.
8. Filter the data to women aged 15--49 and select the final variables
   needed for analysis.
9. Compute the mean and median age.
10. Produce frequency tables for education and residence.
11. Generate a cross-tabulation of education and modern contraceptive use,
    with row percentages.
12. Summarize the number of women and mean age by education level using
    `dplyr`.

---

# Key Takeaways

By the end of this practical, participants should be able to:

- Import and inspect DHS datasets in R
- Clean and prepare data for analysis
- Rename and recode variables
- Handle missing values
- Detect and handle outliers using visual and statistical methods
- Create new variables using `mutate()`
- Filter and select data using `dplyr`
- Produce descriptive statistics and summary tables
